Double White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-07-19 papers/wdmerger_I/paper.tex
Clean up the appendix.

\subsection{Hydrodynamics}

@@

Summary of changes:
\begin{itemize}
\item New gravity update type: fixes some energy conservation

\item changing I's to use limit of the parabola instead of the
  cell-center: not sure if this does much

\item Colella \& Glaz Riemann solver: does a much better job with
  temperature in strong shocks

\item Tracing of gravity: seems to help with the material at the edge
  of the star

\item implicit update of the rotation terms?  (I coded this up once, but
  it is not currently in the code)
\end{itemize}

@@

For all the runs, the PPM reconstruction is done using the original limiters for
the parabolic profiles \citep{ppm}.  The prediction of the interface
states appears as:
\begin{equation}
\label{eq:ppmstatel}
q_{i+1/2,L}^{n+1/2} = \tilde{q}_L -
   \sum_{\nu;\lambda_i^{(\nu)}\ge 0} l_i^{(\nu)} \cdot \left [
        \tilde{q}_L  - \mathcal{I}^{(\nu)}_+(q_i)
       \right ] r_i^{(\nu)}
\end{equation}
where $q$ is the vector of primitive variables, $l_i^{(\nu)}$ and
$r_i^{(\nu)}$ are the left and right eigenvectors with eigenvalue
$\lambda_i^{(\nu)}$, with $\nu$ the index of the characteristic wave of
the system.  The sum is over all the waves that result from the
characteristic structure of the problem, but designed such that only
waves moving toward the interface contribute to the interface value,
$q_{i+1/2,L}^{n+1/2}$.  The reference state, $\tilde{q}_L$ is
chosen to minimize the work of the characteristic projection.
Finally, $\mathcal{I}_+^{(\nu)}(q)$ is the
average under the parabolic profile of quantity $q$ of all the
information that can reach the right interface of the zone $i$ as
carried by the wave $\nu$.  The reader is referred to
\citet{ppmunsplit} for further defaults.

@@

We comment on the choice of reference state for passively-advected
quantities (like $X_k$ or the transverse velocity).
First, consider one of the variables present in one-dimensional flow
(density, velocity in the normal direction, and pressure).  Here we
our reference state is as in Eq.~\ref{eq:refchoice} and we have.
Ignoring flattening, if there are no waves moving toward our
interface, then Eq.~\ref{eq:ppmstatel} reduces to:
\begin{equation}
q_{i+1/2,L}^{n+1/2} = \tilde{q}_L = q_i
\end{equation}
If instead only the fastest wave is moving toward the interface, then
only the term corresponding to the fastest wave in the sum will be
added in Eq.~\ref{eq:ppmstatel}, but our choice of reference state makes
that term 0 by design, and our interface state is:
\begin{equation}
q_{i+1/2,L}^{n+1/2} = \tilde{q}_L = \mathcal{I}_+^{+}(q_i)
\end{equation}
This is the desired behavior for each of these cases.

The details of the reference state for passively advected quantities 
(which includes the transverse velocities in the 1-d tracing) is not
typically discussed.  If we use the same idea of the reference state as
in Eq.~\ref{eq:refchoice}, and consider a quantity $\xi$ which should only
jump across the contact, then our interface state becomes:
\begin{equation}
\xi_{i+1/2,L}^{n+1/2} = \tilde{\xi}_L -
  \underbrace{l_i^\evz \cdot \left [
        \tilde{\xi}_L  - \mathcal{I}^\evz_+(\xi_i)
       \right ] r_i^\evz}_{\text{only if~$u \ge 0$}}
\end{equation}
Again, ignoring flattening, if $u \ge 0$, then we have
\begin{equation}
\xi_{i+1/2,L}^{n+1/2} = \tilde{\xi}_L -
  \left (\tilde{\xi}_L  - \mathcal{I}^\evz_+(\xi_i) \right ) = \mathcal{I}^\evz_+(\xi_i)
\end{equation}
(where we used the fact that the eigenvectors are normalized to 1 and
don't mix in any other states when dealing with passive terms).  This
is the expected behavior---we see a state that is traced only by the
contact wave.  However, if $u < 0$ but $u + c \ge 0$, then we instead get:
\begin{equation}
\xi_{i+1/2,L}^{n+1/2} = \tilde{\xi}_L = \mathcal{I}^\evp_+(\xi_i)
\end{equation}
Here we used the same definition of the reference state and see that our
interface state sees the profile traced under the fastest wave, not the
contact.  This is not the correct behavior for a passively-advected
quantity.

@@

We note a few additional differences between the original PPM
implementation of \citet{ppm} and CASTRO.  In the original PPM
implementation, the gravitational acceleration was reconstructed as a parabola, and
this was traced under to find the forcing that affects the interface
for each wave.  CASTRO followed \citet{ppmunsplit} which instead adds
$(\Delta t/2)g$ to the interface states at the end of the
reconstruction.  In the current implementation, we do the original
parabolic reconstruction and characteristic tracing of the gravitation
source.  Our system with the source appears as:
\begin{equation}
q_t + A(q) q_x = G
\end{equation}
where $G = (0, g, 0)^T$---i.e. the gravitational source only affects
$u$, not $\rho$ or $p$.  Note that in the PPM paper, they put $G$ on 
the lefthand side of the primitive variable equation, so our signs are
opposite.  Our projections are now:
\begin{equation}
\sum_{\nu; \lambda^\enu \ge 0}l^\enu \cdot (\tilde{q} - \mathcal{I}^\enu_+(q) - \tfrac{\Delta t}{2} G) r^\enu
\end{equation}
for the left state, and
\begin{equation}
\sum_{\nu; \lambda^\enu \le 0} l^\enu \cdot (\tilde{q} - \mathcal{I}^\enu_-(q) - \tfrac{\Delta t}{2} G) r^\enu 
\end{equation}
for the right state.  Since $G$ is only non-zero for velocity, only
the velocity changes.  Writing out the sum (and performing the vector products), we
get:
\begin{eqnarray}
u_{i+1/2,L}^{n+1/2} =
   \tilde{u}_+ 
  &-& \frac{1}{2} \left [
      \left (\tilde{u}_+ - \mathcal{I}_+^\evm(u) - \frac{\Delta t}{2} \mathcal{I}^\evm_+(g) \right ) - 
       \frac{\tilde{p}_+ - \mathcal{I}_+^\evm(p)}{C} \right ] \nonumber \\
  &-& \frac{1}{2} \left [
      \left (\tilde{u}_+ - \mathcal{I}_+^\evp(u) - \frac{\Delta t}{2} \mathcal{I}^\evp_+(g) \right ) +
       \frac{\tilde{p}_+ - \mathcal{I}_+^\evp(p)}{C} \right ]
\end{eqnarray}
These differ from the expression in the PPM paper, where $\Delta t G$,
not $\Delta t/2 G$ is used in the projection, however we believe that
the factor of $1/2$ is correct.  To see this, notice that if both
waves are moving toward the interface, then the source term that is
added to the interface state is $(\Delta t/4) (\mathcal{I}_+^\evm(g) +
\mathcal{I}_+^\evp(g))$ for the left state, which reduces to $(\Delta
t/2) g$ for constant g---this matches the result from Taylor
expanding to the interface at the half-time (as in \citealt{ppmunsplit}).

@@

The tracing of gravity can be controlled in the public version of CASTRO
with the parameter {\tt castro.ppm\_trace\_grav}.

@@

\begin{figure}
  \centering
  \includegraphics[scale=1.0]{plots/reference}
  \caption{ \label{Fig:sod} Solution to Sod's problem with the original
    reference state and the new reference state, as compared to the
    exact solution.  We note that the new reference state shows a
    slightly sharper shock and contact, but also has a dip at the tail
    of the rarefaction.}
\end{figure}

>>>>

\subsection{Hydrodynamics}\label{sec:Hydrodynamics}

@@

@@

For all the runs, the PPM reconstruction is done using the original limiters for
the parabolic profiles \citep{ppm}; see Section \ref{sec:Hydrodynamics} for a brief 
discussion about the limiters.  The prediction of the interface
states appears as:
\begin{equation}
\label{eq:ppmstatel}
q_{i+1/2,L}^{n+1/2} = \tilde{q}_L -
   \sum_{\nu;\lambda_i^{(\nu)}\ge 0} l_i^{(\nu)} \cdot \left [
        \tilde{q}_L  - \mathcal{I}^{(\nu)}_+(q_i)
       \right ] r_i^{(\nu)}
\end{equation}
where $q$ is the vector of primitive variables, $l_i^{(\nu)}$ and
$r_i^{(\nu)}$ are the left and right eigenvectors with eigenvalue
$\lambda_i^{(\nu)}$, with $\nu$ the index of the characteristic wave of
the system.  The sum is over all the waves that result from the
characteristic structure of the problem, but designed such that only
waves moving toward the interface contribute to the interface value,
$q_{i+1/2,L}^{n+1/2}$.  The reference state, $\tilde{q}_L$ is
chosen to minimize the work of the characteristic projection.
Finally, $\mathcal{I}_+^{(\nu)}(q)$ is the
average under the parabolic profile of quantity $q$ of all the
information that can reach the right interface of the zone $i$ as
carried by the wave $\nu$.  The reader is referred to
\citet{ppmunsplit} for further details.

@@

We comment on the choice of reference state for passively-advected
quantities (like $X_k$ or the transverse velocity), which is not typically 
discussed. First, consider one of the variables present in one-dimensional flow
(density, velocity in the normal direction, and pressure), and let our 
reference state be as in Eq.~\ref{eq:refchoice}.
Ignoring flattening, if there are no waves moving toward our
interface, then Eq.~\ref{eq:ppmstatel} reduces to:
\begin{equation}
q_{i+1/2,L}^{n+1/2} = \tilde{q}_L = q_i
\end{equation}
If instead only the fastest wave is moving toward the interface, then
only the term corresponding to the fastest wave in the sum will be
added in Eq.~\ref{eq:ppmstatel}, but our choice of reference state makes
that term zero by design, and our interface state is:
\begin{equation}
q_{i+1/2,L}^{n+1/2} = \tilde{q}_L = \mathcal{I}_+^{+}(q_i)
\end{equation}
This is the desired behavior for each of these cases.

However, now consider the same approach applied to passively advected quantities.
If we use the same idea of the reference state as
in Eq.~\ref{eq:refchoice}, and consider a quantity $\xi$ which should only
jump across the contact, then our interface state becomes:
\begin{equation}
\xi_{i+1/2,L}^{n+1/2} = \tilde{\xi}_L -
  \underbrace{l_i^\evz \cdot \left [
        \tilde{\xi}_L  - \mathcal{I}^\evz_+(\xi_i)
       \right ] r_i^\evz}_{\text{only if~$u \ge 0$}}
\end{equation}
Again, ignoring flattening, if $u \ge 0$, then we have
\begin{equation}
\xi_{i+1/2,L}^{n+1/2} = \tilde{\xi}_L -
  \left (\tilde{\xi}_L  - \mathcal{I}^\evz_+(\xi_i) \right ) = \mathcal{I}^\evz_+(\xi_i)
\end{equation}
(where we used the fact that the eigenvectors are normalized to unity and
don't mix in any other states when dealing with passive terms).  This
is the expected behavior---we see a state that is traced only by the
contact wave.  However, if $u < 0$ but $u + c \ge 0$, then we instead get:
\begin{equation}
\xi_{i+1/2,L}^{n+1/2} = \tilde{\xi}_L = \mathcal{I}^\evp_+(\xi_i)
\end{equation}
Here we used the same definition of the reference state and see that our
interface state sees the profile traced under the fastest wave, not the
contact.  This is not the correct behavior for a passively-advected
quantity.

@@

We note a few additional differences between the original PPM
implementation of \citet{ppm} and CASTRO.  In the original PPM
implementation, the gravitational acceleration was reconstructed as a parabola, and
this was traced under to find the forcing that affects the interface
for each wave.  CASTRO originally followed \citet{ppmunsplit} which instead adds
$(\Delta t/2)g$ to the interface states for velocity at the end of the
reconstruction.  In the current implementation, we do the original
parabolic reconstruction and characteristic tracing of the gravitation
source. This can be controlled in CASTRO with the parameter 
{\tt castro.ppm\_trace\_grav}. Our system with the source appears as:
\begin{equation}
q_t + A(q) q_x = G
\end{equation}
where $G = (0, g, 0)^T$---i.e. the gravitational source only affects
$u$, not $\rho$ or $p$.  Note that in the PPM paper, they put $G$ on 
the left-hand side of the primitive variable equation, so our signs are
opposite.  Our projections are now:
\begin{equation}
\sum_{\nu; \lambda^\enu \ge 0}l^\enu \cdot (\tilde{q} - \mathcal{I}^\enu_+(q) - \tfrac{\Delta t}{2} G) r^\enu
\end{equation}
for the left state, and
\begin{equation}
\sum_{\nu; \lambda^\enu \le 0} l^\enu \cdot (\tilde{q} - \mathcal{I}^\enu_-(q) - \tfrac{\Delta t}{2} G) r^\enu 
\end{equation}
for the right state.  Since $G$ is only non-zero for velocity, only
the velocity changes.  Writing out the sum (and performing the vector products), we
get:
\begin{eqnarray}
u_{i+1/2,L}^{n+1/2} =
   \tilde{u}_+ 
  &-& \frac{1}{2} \left [
      \left (\tilde{u}_+ - \mathcal{I}_+^\evm(u) - \frac{\Delta t}{2} \mathcal{I}^\evm_+(g) \right ) - 
       \frac{\tilde{p}_+ - \mathcal{I}_+^\evm(p)}{C} \right ] \nonumber \\
  &-& \frac{1}{2} \left [
      \left (\tilde{u}_+ - \mathcal{I}_+^\evp(u) - \frac{\Delta t}{2} \mathcal{I}^\evp_+(g) \right ) +
       \frac{\tilde{p}_+ - \mathcal{I}_+^\evp(p)}{C} \right ]
\end{eqnarray}
(The expression in the PPM paper contains $\Delta t G$, not $(\Delta t/2) G$,
but we believe that the factor of $1/2$ is correct.  To see this, notice that if both
waves are moving toward the interface, then the source term that is
added to the interface state is $(\Delta t/4) (\mathcal{I}_+^\evm(g) +
\mathcal{I}_+^\evp(g))$ for the left state, which reduces to $(\Delta
t/2) g$ for constant g---this matches the result from Taylor
expanding to the interface at the half-time (as in \citealt{ppmunsplit}).)

@@

@@
